---
title: "Assignment 10: Data Scraping"
author: "Xia Meng Howey"
output: pdf_document
editor_options: 
  chunk_output_type: console
---
## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)


getwd()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
website <- read_html(
  'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')
website
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
systemname <- website %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
systemname

PWSID <- website %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
PWSID

ownership <- website %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

MGD <- website %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
MGD
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
df.Durham.water <- data.frame("Month" = c("Jan", "May", "Sep",
                                          "Feb", "Jun", "Oct",
                                          "Mar", "Jul", "Nov",
                                          "Apr", "Aug", "Dec"),
                              "Year" = rep(2024,12),
                              "System Name" = systemname,
                              "PWSID" = PWSID,
                              "Ownership" = ownership,
                              "MGD" = as.numeric(MGD)) 

df.Durham.water <- df.Durham.water %>% 
  mutate(Date = my(paste(Month, "-", Year))) %>% 
  arrange(Date)

#5 
ggplot(df.Durham.water, aes(
      x=Date,
      y=MGD))+
  geom_line()+
  labs(title = paste("2024 Max Daily Withrawals in Durham"))

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
#Construct the scraping web address
PWSID2 <- '03-32-010'
year2 <- 2024

#create scraping function
scrape.it <- function(PWSID2, year2){
  
  #retrieve website contents
  website.auto <- read_html(paste0(
    'https://www.ncwater.org/WUDC/app/LWSP/report.php','?pwsid=',PWSID2,
    '&year=',year2))
  
  #scrape data items
systemname.auto <- website.auto %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()

PWSID.auto <- website.auto %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()

ownership.auto <- website.auto %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()

MGD.auto <- website.auto %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()

   #convert to a dataframe
df.auto <- data.frame("System Name" = systemname.auto,
                              "PWSID" = PWSID.auto,
                              "Ownership" = ownership.auto,
                              "MGD" = as.numeric(MGD.auto),
                              "Month" = c("Jan", "May", "Sep",
                                          "Feb", "Jun", "Oct",
                                          "Mar", "Jul", "Nov",
                                          "Apr", "Aug", "Dec"),
                              "Year" = rep(year2,12))

df.auto <- df.auto %>% 
  mutate(Date = my(paste(Month, "-", Year))) %>% 
  arrange(Date)

#return dataframe
return(df.auto)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
#use function to scrape Durham data
Durham.auto.df <- scrape.it('03-32-010',2020)

#plot dataframe
ggplot(Durham.auto.df, aes(
  x=Date,
  y=MGD,
))+
  geom_line()+
  labs(title = paste("2020 Max Daily Withrawals in", 
                     Durham.auto.df$System.Name))


#show structure of dataframe
str(Durham.auto.df)
```

8. Use the function above to extract data for Cary (PWSID = '03-32-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
#use function to scrape Cary data
Cary.auto.df <- scrape.it('03-92-020',2020)

#combine Durham and Cary dataframes
Durham.Cary <- bind_rows(Durham.auto.df, Cary.auto.df)
Durham.Cary

#create plot to compare Durham to Cary
ggplot(Durham.Cary, aes(
  x=Date,
  y=MGD,
  colour = System.Name
))+
  geom_line()+
  labs(title = paste("Max Daily Withrawals"))
```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
#Cary max daily withdrawals years 2018-2023
PWSID2 = '03-92-020'
year2 =  2018:2023

the_dfs <- map2(PWSID2, year2,scrape.it) %>% 
  bind_rows()

ggplot(the_dfs,aes(
  x=Date,
  y=MGD
))+
  geom_line()+
  geom_smooth(method="loess",se=FALSE)+
  labs(
    title = paste("Max Daily Withdrawals in Cary")
  )

```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:
>Yes, there is an increase in the max daily withdrawals over time. There also 
>seems to be a seasonal pattern each year.


